Providing a Unified Software Environment for Canada’s National Advanced Computing Centers

### 2019-03-21 paper.tex
Added details about Nix

\subsection{Nix}
\label{sub:Nix}
Nix \cite{Nix} is both a package manager and an operating system (NixOS). One of the core principles in Nix is reproducibility, so it is designed to depend as little as possible on the underlying operating system. Every software package installed through Nix resides in a path calculated with the hash of the recipe to build it. This ensures very strong reproducibility, but it also makes updating a package in place impossible. While reproducibility is an important consideration in advanced research computing, bugs sometimes need to be fixed, which is difficult due to Nix' design. This difficulty arises in particular because our Nix-installed packages are linked into software that is not installed via Nix. At the time of writing this paper, we are still using Nix as a compatibility layer, but we are investigating a switch to Gentoo Prefix.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}
Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer. It allows installation of most packages that form a typical Linux operating system in a non-default location. Contrary to Nix however, packages are installed in a standard directory structure. This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure. At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.

>>>>

\subsection{Nix}
\label{sub:Nix}
Nix~\cite{Nix} is both a package manager and an operating system (NixOS). One of the core principles in Nix is reproducibility, so it is designed to depend as little as possible on the underlying operating system.

The installation path of each software package installed through Nix resides in a path called the Nix store. This stores contains a single directory for each version of each software package installed through Nix. These directories are named after a hash calculated from the installation recipe and its dependencies. This creates a very large number of subdirectories, many of which contain almost identical files.

In addition to the store, Nix creates directories that are called profiles. These profile directories contain a farm of symbolic links that point to files in the Nix store. For example, it will contain a {\it bin} directory, which will contain symbolic links to typical operating system commands such as {\it ls} or {\it gcc}, which will reside in different paths in the store. It will also contain {\it lib} and {\it include} directories which will contain symbolic links to libraries and header files located in the store.

Profiles are the paths that are exposed to the end users, which is done by adding paths to their {\it PATH}, {\it CPATH} and other similar environment variables. Each profile ever created is kept, but users are only exposed to the most recent one. Each profile represents the complete state of the user-exposed environment at a given time in the past.

Nix has a mechanism for garbage collection, which can be used to delete packages that are no longer used in the current profile. It also has a mechanism to resolve the complete list of files from the Nix store that are required to compose a given profile. Because Nix generates so many files, we use this mechanism to deploy on CVMFS only those files that are used in the current profile.

This works well until a software exposes paths to the Nix store directly, instead of the paths to Nix profiles. We call this a Nix store {\it leak}. This happens for example with Python when using virtual environments. Virtual environments in Python copy the {\it python} executable to a secondary location, outside of Nix. However, all binaries installed by Nix use {\it RPATH} for shared object resolutions in the store directly. This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself. Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary. We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed them through EasyBuild instead.

Another issue that we have encountered is that many Nix recipes call for installing sub-packages of a package in separate directories. For example, for Qt, {\it qmake} and {\it OTHER COMMAND} reside in different directories in the Nix store. This sometimes cause issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same {\it bin} folder, alongside with headers in the {\it include} folder, and libraries in the {\it lib} folder. This has mostly been an issue for Qt.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}
Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer. It allows installation of most packages that form a typical Linux operating system in a non-default location. Contrary to Nix however, packages are installed in a standard directory structure ({\it bin}, {\it lib}, {\it include}, {\it share}). This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure. At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.
